### 2022-06-12 paper/paper.tex
paper: final manuscript

% Bibliography
\usepackage[american]{babel}
\usepackage[style=apa, citestyle=apa, backend=biber]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\usepackage[babel,threshold=2]{csquotes}

@@

% Graphs
\usepackage{pgfplots}
\usepgfplotslibrary{external}
\pgfplotsset{width=1\textwidth,compat=1.9}

@@

\begin{center}
	\textbf{Development of a Finger-Key Identification Module\\for a Touch Typing Trainer}
	\par\vspace{0.5cm}
	by
	Oscar Vian L. Valles

@@

Typing tests lists out words for the user to repeat. These tests measure metrics
that quantify the user's ability to type. However, conventional typing metrics
do not measure correct finger placement. This aspect of typing may affect the
user's health. It is also a crucial aspect of keyboard typing education. As
such, a method to identify which finger is used to press which key is
beneficial. This paper introduces a technique to do so by developing a
finger-key identification module that utilizes computer vision algorithms to
detect the keys in a keyboard, and a ready-made machine learning solution to
track fingers while typing. This module was successful in finger-key
identification with an accuracy of 99.47\% in a dataset of 937 keypresses. The
module also had an average finger-key identification time of 0.083 seconds,
which allows the module to perform real-time finger-key identification for
typing speeds up to 143.030 Words per Minute which is well beyond the median
typing speed of the population. Finally, two metrics for correct finger
placement were defined. Finger Placement Accuracy which computes the number of
keys pressed with the correct finger in a typing test sequence. Historical
Finger Placement Accuracy which evaluates the user's ability to press a certain
key with the correct finger over multiple typing test sequences.

@@

Furthermore, Touch typing using correct finger placement is frequently taught in
the beginner level \parencite{donica2018}. This means that there is a need to
weed out bad habits that may develop, like using the index finger for pressing
the spacebar or backspace. However, it is impractical for an educator to check
each student if they are not performing these movements as these may only show
for a small period and may not be caught in time.

@@

The supported keyboard of the module was limited to one specific type of
keyboard. A 60\% Keyboard in a slightly modified \ac{ansi} layout with
light-colored keys as shown in Figure~\ref{fig:60}. This type of keyboard only
has the alphanumeric portion of a full-size keyboard. This limited the number of
keys to be checked and reduced the overall movement of the hand during typing.
The keyboard layout was a slightly modified \ac{ansi} layout. The majority of
the layout still followed the base \ac{ansi} layout, with the alphanumeric
portion remaining exactly the same.

The module's effect on typing was not tested. This includes the module's effect
on typing education and typing health, as the focus of the research was on the
development and testing of the finger-key identification module only.

In addition, the module expected the user to have no deformities in their hands,
and they were capable of typing on a keyboard with difficulties.

@@

\section{Significance of the Research}
This research is beneficial for all users of physical keyboards. These include a
majority of the population as there are a lot of professions that heavily rely
on keyboards. Examples include developers, physicians, educators, and
accountants. By having better ergonomics while typing --- by touch typing
correctly --- wrist injuries can be prevented, and typing speed may be increased

This research also helps educators, especially early educators teaching beginner
typists. By automatically checking for correct technique, the burden of checking
each student is reduced, and directed interventions for bad habits can be
easily created as students with these bad habits are easily identified

@@

Median typing speed using a keyboard with QWERTY layout, based on test data
from \cite{keybr}, is $\approx37.5$ \ac{wpm}. This median is taken from all the test results they have gathered from all of their users.

\cite{monkey-stats} released a dataset showing the speed of 12678 members of the
Monkeytype community with the speed role. The dataset shows the fastest time
that these members have achieved on 60-second tests. The median of the dataset
lies in the 100--109 \ac{wpm} range, with 52.74\% of the community having top
speeds slower than 100 \ac{wpm}.

\cite{arif2009} also performed a survey with 12 participants using
multiple typing methods. They got an average \ac{wpm} of 75.85 with a standard
deviation of 15.61 when their participants used a physical QWERTY layout. This
average \ac{wpm} was also the highest among the other typing methods the
researchers studied.

@@

For those that reach their keys by stretching their fingers, there is an
increased probability that the wrists and fingers adapt to non-neutral postures.
These include wrist extension, ulnar deviation, and pronation, which may cause
musculoskeletal disorders of the upper extremity (\cite{marklin1999} as cited in
~\cite{baker2007})

@@

OpenCV is an open-source computer vision and machine learning software library
that houses ${\approx2500}$ optimized algorithms. This library is widely used by
companies, researchers, and open source communities that utilize computer vision
and machine learning in their projects. Examples of companies that use OpenCV
include Google, Sony, and Honda.

@@

@@

\cite{dorf2001} present a use case for finger tracking in augmented
environments. In the paper, interaction in a virtual environment through the use
of gestures. The tracking system uses an optical marker-based motion capture
system where the user wears a glove with retroreflective markers.

\cite{chiang2014} used a Kinect, a 3D sensing device by Microsoft that
uses depth data, to track fingers to play virtual instruments. Virtual Pianos
and Guitars were created and played with reliable and stable tracking.

\cite{yousaf2014} created a virtual keyboard that operates using finger
tracking. The tracking uses the movement of the finger joints as the basis for
selecting which key to press. A camera captures the movement, and the resulting
video stream is used for hand region detection and finger joint localization.
Using probabilistic regional density-based kernel tracking, finger joint
trajectories are gathered. Feature vectors are then interpreted from the
trajectories. These feature vectors are used in logic-based techniques and
Dynamic Bayesian Network for classification, detection, and recognition of
keystrokes.

@@

@@

The experimental setup and configuration were composed of three elements: the
camera, the keyboard, and the environment. Figure~\ref{fig:metho-setup} is an
image of the experimental setup

@@

\subsection{Camera}
The setup used a single monocular camera positioned in a top-down view. The
camera captured the entirety of the keyboard and the movement of the ten
fingers. To do so, it was mounted on top of the monitor with the camera pointed
down towards the table. Figure~\ref{fig:metho-setup-camera-placement}
illustrates how the camera was mounted over the monitor and
Figure~\ref{fig:metho-setup-keeb} is a picture taken with the camera at this
position.

@@

The specific camera used was a Logitech C920. It is a 3 mega pixel webcam that
is capable of capturing color video in 1080p/30fps and 720p/30fps with a
diagonal field of view of 78\degree. This camera has a universal mounting clip
that allows the camera to be correctly positioned within the experimental setup
\parencite{logitech}.

@@

In addition, the keyboard layout was a modified \ac{ansi} layout. The majority
of the keys remained the same, however some keys were changed: The pipe key was
replaced by the backspace key, the backspace key was replaced by the pipe key
and a delete key, and the bottom row has a 7u length spacebar. Overall, these
are minor changes that does not affect the alphanumeric keys that are
predominantly used in typing.

\subsection{Computer}
The computer used to run the algorithm is a desktop computer with the following
specification:

\begin{table}[H]
	\small
	\centering
	\begin{tabular}{ p{0.3\textwidth} p{0.6\textwidth} }
		\toprule
		Component & Specification                          \\
		\midrule
		CPU       & AMD Ryzen 5 3600                       \\
		GPU       & AMD Radeon RX 5600 XT                  \\
		RAM       & G.Skill Trident Z Neo RGB 16GB 3200mhz \\
		Storage   & Samsung SSD 850 EVO                    \\
		OS        & Arch Linux                             \\
		\bottomrule
	\end{tabular}
	\caption{\label{tab:table-specs}Computer Specifications}
\end{table}

\subsection{Environment}
The setup was lit with a lamp besides the camera. This light source evenly lit
the keyboard and the fingers used for typing. The light source used was a common
LED bulb rated at 9 Watts with 700 lumens. This light was white with a color
temperature of 6500k.

@@

\paragraph{Edge Detection}
The algorithm used a Sobel filter for edge detection. This filter uses a
3$\times$3 kernel that is convolved twice. Once horizontally, and another
vertically to produce a grayscale image of the outlines within the frame. The
kernels used by the Sobel filter \parencite{sobel2014} is shown in
Figure~\ref{fig:metho-algo-key-sobel} and Figure~\ref{fig:metho-algo-sobel} is
the output of this step.

@@

\paragraph{Thresholding}
The output of the Sobel filter is a grayscale image of 255 values, with each
gray pixel indicating edges within the image. There is a need to reduce the
range of these values to improve the ability of the algorithm to find the
contours of these edges. To do so, the algorithm utilized Otsu's algorithm to
perform automated thresholding. Otsu's algorithm determines a single threshold
that is most optimal for the image \parencite{otsu}. This outputs a
black-and-white image, with only two values. Figure~\ref{fig:metho-algo-thresh}
is the resulting output of this step.

@@

\paragraph{Find Contours}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. The OpenCV function
\texttt{findContours}, with the \texttt{CHAIN\_APPROX\_SIMPLE} contour
approximation method, was used to find the contours of the outlines of the
object found using the previous step. This approximation method removed
redundant points and returned the least amount of points that describe the
shape. This OpenCV function implements the algorithm of~\cite{contours} in their
paper \citetitle{contours}. Figure~\ref{fig:metho-algo-contours} shows the contours
found in the threshed image.

@@

The contour points that described the contour were then sorted clockwise. This
was a necessity since the algorithm of~\cite{contours} does not guarantee that
the largest contour's points were returned in a clockwise arrangement, which is
a requirement for the next step. Figure~\ref{fig:metho-algo-largest-contour}
shows the four contour points found at the edges of the keyboard within the
image.

@@

\paragraph{Transform Image Map}
The virtual keyboard map is a rectangular image that contains individual
\ac{roi} for each key in a 60\% \ac{ansi} keyboard. Each key has a corresponding
color assigned to it and this color fills the region where this key is located
at. Figure~\ref{fig:metho-algo-key-map} is the initial image map and Appendix
Chapter~\ref{appendix:key-map} is a table that shows the hex color code and key
mapping present in the image map.

@@

\paragraph{Scale Contours}
During testing using training data, a lot of the failures in finger-key
identification were due to the tight fit of the contour to the key. The contour
on its own did not give enough buffer for the placement of the finger. This
buffer was needed as, in some instances, the finger used to press the key was
not exactly on top of the key, but rather to its side. The algorithm accounts
for this situation by adding a buffer. This was achieved by scaling the contour
points by seven pixels on each side. Figure~\ref{fig:metho-algo-scale-roi} shows
the increased buffer in relation to the key. This value was obtained after
testing other possible values. A buffer of five pixels did not have a lot of
effect in reducing the number of failures. A buffer of ten pixels did reduce the
number of failures, but it also greatly increased the algorithm's uncertainty in
determining the finger --- uncertainty is defined as detecting two or more
fingers within one \ac{roi}. Seven was a good middle ground in decreasing
failures, without greatly increasing uncertainty.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{scaled-contour.png}
	\caption{Contour points of the \ac{roi}}
	\label{fig:metho-algo-scale-roi}
	\centering
\end{figure}

@@

\subsubsection{Palm Detection Model}
The first model, the Palm Detection Model detects the initial hand locations
using a single shot detector model based on the paper by~\cite{ssd}. This model
achieves an average precision of 95.7\% in palm detection
\parencite{mediapipe-hands}.

@@

\subsection{Implementation}
\label{section:metho-algo-implementation}
The previously discussed algorithms were implemented using Python. The Keyboard
Detection and Mapping Algorithm in Section~\ref{section:metho-algo-keyboard} was
implemented as a separate module in Python using OpenCV. The Finger Detection
and Tracking Solution in Section~\ref{section:metho-algo-finger} was consumed
using the prebuilt Python package offered by the MediaPipe team. Finally, the
integration of the two will be done as another separate module in Python.

@@

There was a total of 10 test sequences that were gathered. An example test
sequence is ``What of it, if some old hunks of a sea-captain orders me to get a
broom and sweep down the decks?'' from Moby Dick by~\cite{moby-dick}. The 9
other test sequences can be found in Appendix
Chapter~\ref{appendix:test-sequences}.

\subsection{Data Gathering}
\subsubsection{Video Capture}
There were 3 groups of 10 videos that were captured by the researcher, for a
total of 30 videos. Each group corresponds to three predetermined speeds of
typing: slow (15wpm), average (35wpm), and fast (80wpm). These speeds were based
on test data from~\cite{keybr}. The researcher typed the 10 typing test
sequences at these predetermined speeds. During typing, errors were not
consciously taken into account, and errors happened naturally as part of the
typing process.

@@

\subsubsection{Keypress Labeling}
The researcher then manually perform finger-key identification for all key
presses present in the video. This was also done through a Python script. The
script parsed the file associated with a video. The frame number found at each
line was then shown, and the associated data with the frame was superimposed
over it. Figure~\ref{fig:metho-data-process} is an example of a frame with these
data overlaid over the captured image. The researcher then pressed keys that
corresponded with the finger used to press the key. This was then again stored
in the format \texttt{FRAME\_NO:KEY\_PRESSED:FINGER\_USED}, and each keypress
was then stored as a line in a file.

@@

@@

\begin{table}[H]
	\small
	\centering
	\begin{tabular}{ p{0.3\textwidth} p{0.6\textwidth} }
		\toprule
		Column Name                           & Definition                                             \\
		\midrule
		\texttt{file\_name}                   & Name of the file                                       \\
		\texttt{frame}                        & Frame number                                           \\
		\texttt{key}                          & Key pressed                                            \\
		\texttt{finger}                       & Manual labeling of the finger used to press the key    \\
		\texttt{is\_correct}                  & If the module was correct                              \\
		\texttt{fingertips\_detected}         & The fingertips detected by the module                  \\
		\texttt{keyboard\_detection\_time}    & Time spent in creating the Key-Edge Coordinates Map    \\
		\texttt{finger\_classification\_time} & Time spent in classifying which finger pressed the key \\
		\bottomrule
	\end{tabular}
	\caption{\label{tab:csv-output}CSV File Contents}
\end{table}

Adjustments were then made to the algorithm after each training pass to improve
its performance. Some runs had improvements in its statistics without doing any
adjustments. These were runs that showed errors in the manual labeling and
were corrected afterward.

\subsection{Analysis}
The resulting \texttt{.csv} file was imported into Google Sheets and key
statistics were obtained based on the data. These statistics can be found on
Table~\ref{tab:acquired-stats}. These key statistics informed what to adjust in
the algorithm and its overall performance.

>>>>

% Bibliography
\usepackage[american]{babel}
\usepackage[style=apa, citestyle=apa, sortcites=true, backend=biber]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\usepackage[babel,threshold=2]{csquotes}

@@

% Graphs
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepgfplotslibrary{external}
\pgfplotsset{width=1\textwidth,compat=1.9}

@@

\begin{center}
	\textbf{Development of a Finger-Key Identification Module\\for a Touch Typing Trainer}\\
	by\\
	Oscar Vian L. Valles

	\par\vspace{0.5cm}

@@

Typing tests list out words for the user to repeat. These tests measure metrics
that quantify the user's ability to type. However, conventional typing metrics
do not measure correct finger placement. This aspect of typing may affect the
user's health. It is also a crucial aspect of keyboard typing education. As
such, a method to identify which finger is used to press which key is
beneficial. This paper introduces a technique to achieve this by developing a
finger-key identification module that utilizes computer vision algorithms to
detect the keys in a keyboard, and a ready-made machine learning solution to
track fingers while typing. This module was successful in finger-key
identification with an accuracy of 99.58\% in a dataset of 938 keypresses. The
module also had an average finger-key identification time of 0.083 seconds,
which allows the module to perform real-time finger-key identification for
typing speeds up to 143.030 Words per Minute which is well beyond the median
typing speed of the population. Furthermore, two metrics for correct finger
placement were defined and were implemented in a proof-of-concept trainer: (1)
Finger Placement Accuracy which computes the number of keys pressed using the
correct finger in a typing test sequence, and (2) Historical Finger Placement
Accuracy which evaluates the user's ability to press a certain key with the
correct finger over multiple typing test sequences.

@@

Furthermore, touch typing using correct finger placement is frequently taught in
the beginner level \parencite{donica2018}. This means that there is a need to
weed out bad habits that may develop, like using the index finger for pressing
the spacebar or backspace. However, it is impractical for an educator to check
each student if they are not performing these movements as these may only show
for a small period and may not be caught in time.

@@

The supported keyboard of the module was limited to one specific type of
keyboard. A 60\% Keyboard in a slightly modified \ac{ansi} layout with
light-colored keys was used as shown in Figure~\ref{fig:60}. This type of
keyboard only has the alphanumeric portion of a full-size keyboard. This limited
the number of keys to be checked and reduced the overall movement of the hand
during typing. The keyboard layout was a slightly modified \ac{ansi} layout. The
majority of the layout still followed the base \ac{ansi} layout, with the
alphanumeric portion remaining exactly the same.

The module's effect on typing was not tested. This includes the module's effect
on typing in education and typing in health, as the focus of the research was on
the development and testing of the finger-key identification module only.

@@

\section{Significance of the Research}
This research is beneficial for all users of physical keyboards. These include a
majority of the population as there are a lot of professions that heavily rely
on keyboards. Examples include developers, physicians, educators, and
accountants. By having better ergonomics while typing --- by touch typing
correctly --- wrist injuries can be prevented, and typing speed may be
increased.

This research also helps educators, especially early educators teaching beginner
typists. By automatically checking for correct technique, the burden of checking
each student is reduced, and directed interventions for bad habits can be easily
created as students with these bad habits are easily identified.

@@

Median typing speed using a keyboard with QWERTY layout, based on test data from
\textcite{keybr}, is $\approx37.5$ \ac{wpm}. This median is taken from all the test
results they have gathered from all of their users.

\textcite{monkey-stats} released a dataset showing the speed of 12678 members of the
Monkeytype community with the speed role. The dataset shows the fastest time
that these members have achieved on 60-second tests. The median of the dataset
lies in the 100--109 \ac{wpm} range, with 52.74\% of the community having top
speeds slower than 100 \ac{wpm}.

\textcite{arif2009} also performed a survey with 12 participants using
multiple typing methods. They got an average \ac{wpm} of 75.85 with a standard
deviation of 15.61 when their participants used a physical QWERTY layout. This
average \ac{wpm} was also the highest among the other typing methods the
researchers studied.

@@

For those that reach their keys by stretching their fingers, there is an
increased probability that the wrists and fingers adapt to non-neutral postures.
These include wrist extension, ulnar deviation, and pronation, which may cause
musculoskeletal disorders of the upper extremity (\textcite{marklin1999} as cited in
~\textcite{baker2007})

@@

OpenCV is an open-source library that provides ${\approx2500}$ optimized
algorithms for computer vision and machine learning. This library is widely used
by companies, researchers, and open source communities that utilize computer
vision and machine learning in their projects. Examples of companies that use
OpenCV include Google, Sony, and Honda.

@@

Examples of projects that utilized MediaPipe and its solutions include a
real-time hand gesture recognition system by \textcite{mediapipe-sample-1} which
utilized MediaPipe Hands. Another example is by \textcite{mediapipe-sample-2}
which utilized MediaPipe Hands to track movements during pen spinning
competitions. \textcite{mediapipe-sample-3} developed a real-time sign language
recognition program using MediaPipe Hands.

@@

\textcite{dorf2001} present a use case for finger tracking in augmented
environments. In the paper, interaction in a virtual environment through the use
of gestures. The tracking system uses an optical marker-based motion capture
system where the user wears a glove with retroreflective markers.

\textcite{chiang2014} used a Kinect, a 3D sensing device by Microsoft that
uses depth data, to track fingers to play virtual instruments. Virtual Pianos
and Guitars were created and played with reliable and stable tracking.

\textcite{yousaf2014} created a virtual keyboard that operates using finger
tracking. The tracking uses the movement of the finger joints as the basis for
selecting which key to press. The tracking first detects where the hands are.
The result of the detection is then piped to a different algorithm for finger
joint localization. These are then used to generate finger joint trajectories
using kernel tracking. A Dynamic Bayesian Network then uses feature vectors from
these trajectories to classify, detect, and recognize keystrokes.

@@

This section provides an overview of the development and testing of a real-time
finger-key identification module. First, the experimental setup used for data
collection will be presented. Following that, the algorithm used to implement
real-time finger-key identification will be discussed. The process of training
and testing of the developed algorithm comes next. Afterwards, the development
and design of the proof of concept trainer will be described. Finally, this
section introduces two metrics that were developed to quantify correct finger
placement during typing tests.

@@

The experimental setup and configuration were composed of three elements: the
camera, the keyboard, and the environment. This setup was used to collect data
points to measure the algorithm's ability to perform finger-key identification
in a controlled environment. Figure~\ref{fig:metho-setup} is an image of the
experimental setup.

@@

\subsection{Camera}
The setup used a single monocular camera positioned in a top-down view. The
camera captures the entirety of the keyboard and the movement of the ten
fingers. To do so, it was mounted on top of the monitor with the camera pointed
down towards the table. Figure~\ref{fig:metho-setup-camera-placement}
illustrates how the camera was mounted over the monitor and
Figure~\ref{fig:metho-setup-keeb} is a picture taken with the camera at this
position.

@@

The specific camera used was a Logitech C920. It is a 3 mega pixel webcam that
is capable of capturing color video in 1080p/30fps and 720p/30fps with a
diagonal field of view of 78\degree. For the setup, the camera captured the
video in 480p/30fps. This camera has a universal mounting clip that allows the
camera to be correctly positioned within the experimental setup
\parencite{logitech}.

@@

In addition, the keyboard layout was a modified \ac{ansi} layout. The majority
of the keys remained the same, however some keys were changed: the pipe key was
replaced by the backspace key, the backspace key was replaced by the pipe key
and a delete key, and the bottom row has a longer spacebar. Overall, these are
minor changes that does not affect the alphanumeric keys that are predominantly
used in typing.

\subsection{Computer}
The computer used to run the algorithm is a desktop computer. The specifications
of this computer is shown in Table~\ref{tab:table-specs}.

\begin{table}[H]
	\small
	\centering
	\caption{\label{tab:table-specs}Computer Specifications}
	\begin{tabular}{ p{0.3\textwidth} p{0.6\textwidth} }
		\toprule
		Component & Specification                          \\
		\midrule
		CPU       & AMD Ryzen 5 3600                       \\
		GPU       & AMD Radeon RX 5600 XT                  \\
		RAM       & G.Skill Trident Z Neo RGB 16GB 3200mhz \\
		Storage   & Samsung SSD 850 EVO                    \\
		OS        & Arch Linux                             \\
		\bottomrule
	\end{tabular}
\end{table}

\subsection{Environment}
The setup was lit with a lamp besides the camera. This light source evenly lit
the keyboard and the fingers used for typing. The light source used was a common
LED bulb rated at 9 watts with 700 lumens. This light is white with a color
temperature of 6500k.

@@

\paragraph{Edge Detection}
The algorithm used a Sobel filter for edge detection. This filter uses a
3$\times$3 kernel that is convolved twice --- once horizontally, and another
vertically --- to produce a grayscale image of the outlines within the frame.
The kernels used by the Sobel filter \parencite{sobel2014} is shown in
Figure~\ref{fig:metho-algo-key-sobel} and the output of this step is shown in
Figure~\ref{fig:metho-algo-sobel}.

@@

\paragraph{Thresholding}
The output of the Sobel filter is a grayscale image of 256 values, with each
gray pixel indicating edges within the image. There is a need to reduce the
range of these values to improve the ability of the algorithm to find the
contours of these edges. To do so, the algorithm utilized Otsu's algorithm to
perform automated thresholding. Otsu's algorithm determines a single threshold
that is most optimal for the image \parencite{otsu}. This outputs a
black-and-white image, with only two values. Figure~\ref{fig:metho-algo-thresh}
is the resulting output of this step.

@@

\paragraph{Find Contours}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. The OpenCV function
\texttt{findContours}, with the \texttt{CHAIN\_APPROX\_SIMPLE} contour
approximation method, was used to find the contours of the outlines of the
object found using the previous step. This approximation method removed
redundant points and returned the least amount of points that describe the
shape. This OpenCV function implements the algorithm of~\textcite{contours} in their
paper \citetitle{contours}. Figure~\ref{fig:metho-algo-contours} shows the contours
found in the threshed image.

@@

The contour points that described the contour were then sorted clockwise. This
was a necessity since the algorithm of~\textcite{contours} does not guarantee
that the largest contour's points were returned in a clockwise arrangement.
There is a need to have consistent ordering of these points since the
transformation that will be performed in the next step pairs the contour points
to a hard-coded array that indicates the edges of the image map. This hard-coded
array was sorted clockwise. As such, if the contour points were not sorted
clockwise, there will be no guarantee that the pairing would result in a
transformation that creates a transformed image map that is oriented correctly,
since the pairing may not pair the correct points.

Figure~\ref{fig:metho-algo-largest-contour} shows the four contour points found
at the edges of the keyboard within the image.

@@

\paragraph{Transform Image Map}
The virtual keyboard map is a rectangular image that contains individual
\ac{roi} for each key in a 60\% \ac{ansi} keyboard. Each key has a corresponding
color assigned to it and this color fills the region where this key is located
at. Figure~\ref{fig:metho-algo-key-map} is the initial image map and
Appendix~\ref{appendix:key-map} is a table that lays out the mapping of the
color code in BGR format and its associated character in the image map.

@@

\paragraph{Scale Contours}
During testing using training data, a lot of the failures in finger-key
identification were due to the tight fit of the contour to the key. The contour
on its own did not give enough buffer for the placement of the finger. This
buffer was needed as, in some instances, the finger used to press the key was
not exactly on top of the key, but rather to its side. In addition, it can also
be observed that the finger tracking does not annotate the very edge of the
fingertip, but rather, it labels somewhere in the middle of the fingernail.

The algorithm accounts for this situation by adding a buffer. This was achieved
by scaling the contour points by seven pixels on each side.
Figure~\ref{fig:metho-algo-scale-roi} shows the increased buffer in relation to
the key. This value was obtained after testing other possible values. A buffer
of five pixels did not have a lot of effect in reducing the number of failures.
A buffer of ten pixels did reduce the number of failures, but it also greatly
increased the algorithm's uncertainty in determining the finger --- uncertainty
is defined as detecting two or more fingers within one \ac{roi}. Seven was a
good middle ground in decreasing failures, without greatly increasing
uncertainty.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{scaled-contour.png}
	\caption{Scaled contour points of the \ac{roi} with the buffer of seven pixels}
	\label{fig:metho-algo-scale-roi}
	\centering
\end{figure}

@@

\subsubsection{Palm Detection Model}
The first model, the Palm Detection Model detects the initial hand locations
using a single shot detector model based on the paper by~\textcite{ssd}. This model
achieves an average precision of 95.7\% in palm detection
\parencite{mediapipe-hands}.

@@

\subsection{Implementation}
\label{section:metho-algo-implementation}
The previously discussed algorithms were implemented using Python. The Keyboard
Detection and Mapping Algorithm in Section~\ref{section:metho-algo-keyboard} was
implemented as a separate module in Python using OpenCV. The Finger Detection
and Tracking Solution in Section~\ref{section:metho-algo-finger} was implemented
using the prebuilt Python package offered by the MediaPipe team. Finally, the
integration of the two was done as another separate module in Python.

@@

There was a total of 10 test sequences that were gathered. An example test
sequence is ``What of it, if some old hunks of a sea-captain orders me to get a
broom and sweep down the decks?'' from Moby Dick by~\textcite{moby-dick}. The 9
other test sequences can be found in Appendix
Chapter~\ref{appendix:test-sequences}.

\subsection{Data Gathering}
\subsubsection{Video Capture}
There were 3 sets of 10 videos that were captured by the researcher, for a total
of 30 videos. Each set corresponds to three predetermined speeds of typing: slow
(15wpm), average (35wpm), and fast (80wpm). These speeds were based on test data
from~\textcite{keybr}. The researcher typed the 10 typing test sequences at
these predetermined speeds. Best effort was made to stick to the predetermined
speeds. During typing, errors were not consciously taken into account, and
errors happened naturally as part of the typing process --- with more errors
happening more frequently as the typing speed increased.

@@

\subsubsection{Keypress Labeling}
The researcher then manually performed finger-key identification for all key
presses present in the video. This was also done through a Python script. The
script parsed the file associated with a video. The frame number found at each
line was then shown, and the associated data with the frame was superimposed
over it. Figure~\ref{fig:metho-data-process} is an example of a frame with these
data overlaid over the captured image. The researcher then pressed keys that
corresponded with the finger used to press the key. This was then again stored
in the format \texttt{FRAME\_NO:KEY\_PRESSED:FINGER\_USED}, and each keypress
was then stored as a line in a file.

@@

\subsubsection{Data Inventory}

\begin{figure}[H]
	\centering
	\begin{tikzpicture}
		\begin{axis}[
				xbar, xmin=0,
				bar width=4pt,
				width=0.8\textwidth,
				ylabel=Key,
				xlabel=Count,
				symbolic y coords={
						J,
						Q,
						Dash,
						ForwardSlash,
						Z,
						1,
						LeftAlt,
						X,
						Comma,
						K,
						V,
						G,
						B,
						Backspace,
						W,
						Period,
						P,
						U,
						C,
						LeftShift,
						F,
						Y,
						M,
						H,
						D,
						L,
						I,
						R,
						N,
						T,
						S,
						A,
						O,
						E,
						Spacebar,
					},
				ytick=data,
				tick label style={font=\footnotesize},
				label style={font=\footnotesize},
				xmajorgrids=true
			]
			\addplot table[x=Value,y=Key]{training-data.txt};
		\end{axis}
	\end{tikzpicture}
	\caption{Training Dataset Inventory}
	\label{fig:metho-inventory-training}
\end{figure}

Individual counts of the characters in the training iteration data set is shown
in Figure~\ref{fig:metho-inventory-training}. The letters were not represented
equally in the data set. Of note, J and Q were not in the dataset.

\begin{figure}[H]
	\centering
	\begin{tikzpicture}
		\begin{axis}[
				xbar, xmin=0,
				bar width=4pt,
				width=0.8\textwidth,
				ylabel=Key,
				xlabel=Count,
				symbolic y coords={
						J,
						Q,
						Z,
						Dash,
						ForwardSlash,
						X,
						V,
						Period,
						K,
						B,
						Comma,
						G,
						F,
						Backspace,
						P,
						W,
						C,
						U,
						M,
						Y,
						D,
						LeftShift,
						L,
						H,
						N,
						R,
						S,
						I,
						O,
						A,
						T,
						E,
						Spacebar,
					},
				ytick=data,
				tick label style={font=\footnotesize},
				label style={font=\footnotesize},
				xmajorgrids=true
			]
			\addplot table[x=Value,y=Key]{test-data.txt};
		\end{axis}
	\end{tikzpicture}
	\caption{Test Dataset Inventory}
	\label{fig:metho-inventory-test}
\end{figure}

The letters were also not represented equally in the test dataset as shown in
Figure~\ref{fig:metho-inventory-test}. Three characters in the alphabet were not
represented in the dataset: J, Q, and Z. The inequality of representation in
both of these datasets can be attributed to the source materials of the typing
test sequences --- novels and texts. While there were criteria to select which
text was sourced, there was no conscious decision to equally represent all
characters in the data set.

@@

\begin{table}[H]
	\small
	\centering
	\caption{\label{tab:csv-output}CSV File Contents}
	\begin{tabular}{ p{0.3\textwidth} p{0.6\textwidth} }
		\toprule
		Column Name                           & Definition                                             \\
		\midrule
		\texttt{file\_name}                   & Name of the file                                       \\
		\texttt{frame}                        & Frame number                                           \\
		\texttt{key}                          & Key pressed                                            \\
		\texttt{finger}                       & Manual labeling of the finger used to press the key    \\
		\texttt{is\_correct}                  & If the module was correct                              \\
		\texttt{fingertips\_detected}         & The fingertips detected by the module                  \\
		\texttt{keyboard\_detection\_time}    & Time spent in creating the Key-Edge Coordinates Map    \\
		\texttt{finger\_classification\_time} & Time spent in classifying which finger pressed the key \\
		\bottomrule
	\end{tabular}
\end{table}

Adjustments were then made to the algorithm after each training pass to improve
its performance. Some runs had improvements in their statistics without doing
any adjustments. These were runs that showed errors in the manual labeling and
were corrected afterward.

\subsection{Analysis}
The resulting \texttt{.csv} file was imported into Google Sheets and key
statistics were obtained based on the data. Appendix~\ref{appendix:sheets} shows
the screenshots where this was performed. The statistics obtained can be found
in Table~\ref{tab:acquired-stats}. These key statistics informed what to adjust
in the algorithm and its overall performance.

\begin{table}[H]
	\small
	\centering
	\caption{\label{tab:acquired-stats}Acquired statistics}
	\begin{tabular}{ p{0.3\textwidth} p{0.6\textwidth} }
		\toprule
		Identifications                                                                                                                    \\
		\midrule
		Successes                                & Number of successful finger-key identifications                                         \\
		Failures                                 & Number of failed finger-key identifications                                             \\
		Success Percentage                       & Percentage of successful finger-key identifications                                     \\
		Failure Percentage                       & Percentage of failed finger-key identifications
		\\[0.25cm]
		\midrule
		Failure Types                                                                                                                      \\
		\midrule
		Mismatched Identification                & Module output did not match manual labeling                                             \\
		No Identification                        & Module did not detect fingertips in \ac{roi}
		\\[0.25cm]
		\midrule
		Uncertain Successes                                                                                                                \\
		\midrule
		Total                                    & Total uncertain successes, defined as detecting two or more fingers within one \ac{roi} \\
		Spacebar                                 & All uncertain successes that were on the spacebar                                       \\
		Spacebar Percentage                      & Percentage of spacebar uncertain successes over all uncertain successes                 \\
		Spacebar Percentage (Total)              & Percentage of spacebar uncertain successes over total keypresses                        \\
		Non-spacebar                             & All other uncertain successes                                                           \\
		Non-spacebar Percentage                  & Percentage of non-spacebar uncertain successes over all uncertain successes             \\
		Non-spacebar Percentage \newline (Total) & Percentage of non-spacebar uncertain successes over total keypresses
		\\[0.25cm]
		\midrule
		Average Running Time                                                                                                               \\
		\midrule
		Keyboard Detection Time                  & Average keyboard detection time of all videos                                           \\
		Finger Identification Time               & Average finger identification time of all key presses                                   \\
		\bottomrule
	\end{tabular}
\end{table}
